---
title: "World population data"
author: "John Nesemann, Kevin Tang, Ian McCormick"
date: "9/8/2021"
output: 
  html_document:
   toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up

Load packages
```{r packages}
# remotes::install_github('wpgp/wopr') # you first need to install the packages directly from github
# library(wopr)
library(tidyverse)
library(here)
library(sf)
library(wpgpDownloadR)
library(raster)
library(sp)
library(rgdal)
library(viridis)
library(tmap)
```
Load data
```{r data}
eye_fac <- read_csv(here("data","Malawi_facilities_gps.csv")) %>%
  rename(cat_serv=`Cataract surgical services`)
eye_fac_sf <- eye_fac %>% 
  st_as_sf(coords = c("Long","Lat"), # converting to an sf object
           crs = "4326") # setting crs to be same as that used by world pop (WGS84)
malawi <- st_read(here("data","mwi_adm_nso_20181016_shp.nosync","mwi_admbnda_adm2_nso_20181016.shp"))

```

## Download population data.

World pop has two main methods of top down estimation modelling population data: constrained and unconstrained (https://www.worldpop.org/methods/top_down_constrained_vs_unconstrained). Essentially unconstrained estimates population over all land grid squares whereas constrained estimates only within areas mapped as containing built settlements. For the rest of this analysis we will use the constrained estimates as these are less likely to predict small population numbers in likely uninhabited areas; however, these estimates are only available for 2020. We could also consider using unconstrained estimates as these may produce better looking maps and then do sensitivity analyses with constrained estimates.

The data is only available in raster form and is further divided by age group. Thus we will first need to download all the rasters we are interested in, merge them, and then extract meaningful population data to map.

Note: Knitr does not like this section of code, probably because it takes a very long time. I am inactivating it for knitr but you can reactivate it independently to download the data to your own computer.

```{r download world pop data}
wpgpListCountries() %>%                  # lists all ISO3 (country code) available on WorldPop ftp server
   filter(Country=="Malawi")              # confirms Malawi is available and has ISO3 = MWI

wpgpListCountryDatasets(ISO3 = "MWI")    # returns dataframe of available covariates for Malawi

# The data can be downloaded as raster files
wpgpGetCountryDataset(ISO3 = "MWI",   # ISO3 of desired country
                     covariate = "agesex_m_50_2020_constrained", # covariate(s) desired
                     destDir = "~/Library/Mobile Documents/com~apple~CloudDocs/R projects/eye_facilities/data/pop_constrained") # directory on computer to download the files

# However it will only download one age grouping (e.g., females 50-54) at a time

# creating a list of covariates we want to download
  raster.list <- wpgpListCountryDatasets(ISO3 = "MWI") %>%
    # filtering out the specific covariates we want (constrained estimates of age and sex distribution for those 50+)
    filter(grepl("agesex_", Covariate)) %>%
    filter(grepl("_50", Covariate) | grepl("_55", Covariate) | grepl("_60", Covariate) |
             grepl("_65", Covariate) | grepl("_70", Covariate) | grepl("_75", Covariate) |
             grepl("_80", Covariate)) %>%
    filter(grepl("_2020_constrained$", Covariate)) %>%
    dplyr::select(Covariate)
  raster.list

  # turning that list into a vector
  raster.list.v <- raster.list$Covariate

# Building a function to download all of the variables of interest

get.raster <- function(x, dir) {

  # setting ISO
  ISO3 <- "MWI"
  # setting destination directory
  destDir <- dir
  # getting a raster
  raster <- wpgpGetCountryDataset(ISO3 = ISO3,
                                covariate = x,
                                destDir = destDir)
  # returning the results
  raster

}

# testing the function
get.raster("agesex_f_55_2020_constrained", here("data", "pop_constrained"))   # works for a single covariate

# applying it over all the desired covariates
sapply(X = raster.list.v,  # the covariates we want rasters for
       dir = here("data", "pop_constrained"),
       FUN = get.raster,   # the function to apply
       simplify = T)
```

## Stack the individual age and sex group rasters and combine them into a single raster.

Each raster file had only the estimates for it's corresponding age and sex band, what we want to do now is combine them all into a multi-layered raster AKA a stack. Once we have this stack we can combine all the layers to generate plot the number of individuals 50+ per square km alongside the cataract surgical centers. We can also extract the data into a dataframe which can then be used for analysis.

Note: Knitr does not like this section of code, probably because it takes a very long time. I am inactivating it for knitr but you can reactivate it independently to download the data to your own computer.

```{r combining the individual world pop rasters}
#### first I will check my work to ensure the functions are doing what I desire. ####
# loading raster into an R object
m50to54 <- raster(here("data", "pop_constrained", "mwi_agesex_m_50_2020_constrained.tif"))
names(m50to54) <- "m_50" # look at the raster attributes, it is a RasterLayer
m55to59 <- raster(here("data", "pop_constrained", "mwi_agesex_m_55_2020_constrained.tif"))
names(m55to59) <- "m_55"

# creating a stack
ts.stack <- stack(m50to54, m55to59)
ts.stack

# calculating new raster which is the sum of each raster in the stack
ts.sum <- calc(ts.stack, fun=sum)
ts.sum
summary(ts.sum)

# repeating the process on dataframes to check my work

  # converting to dataframes
  m_50df <- as.data.frame(rasterToPoints(m50to54))
  m_55df <- as.data.frame(rasterToPoints(m55to59))

  # joining the two
  mmerge <- full_join(m_50df, m_55df,  by = c("x", "y")) %>%
    # calculating total values for each point
    mutate(total=m_50+m_55)

  # summarizing
  summary(mmerge$total)
  # and comparing
  summary(ts.sum)
  # looks like it works
  
#### repeating the process for all of the downloaded rasters ####

  # getting a list of all filenames
  list <- list.files(here("data", "pop_constrained"), full.names = T)

  # creating the stack
  popstack <- stack(list)
  popstack

  # giving the stack more intelligible names
  names(popstack) <- c("f_50","f_55","f_60","f_65","f_70", "f_75", "f_80",
                     "m_50","m_55","m_60","m_65","m_70", "m_75", "m_80")

  # calculating a new raster which is the sum of the population values in all the stacks
  popsum <- calc(popstack, fun=sum)

  # renaming the newly created layer
  names(popsum) <- "total_pop"

  # saving the newly created raster
  writeRaster(popsum, here("data", "rasters", "popsum_constrained.tif"))
```

## Plot the resulting raster with permanent cataract surgical facilities.

The only package I have found that allows plotting of raster and sf objects together is the tmap package.
```{r mapping using constrained data}

# Opening the tiff file created above (can do this instead of downloading and mergin everything everytime you run the code)
popsum <- raster(here("data", "rasters", "popsum_constrained.tif"))

# plotting with tmap
constrained_map <- tm_shape(malawi) + tm_polygons() +
  tm_shape(popsum) +
  tm_raster(style = "order",
            palette = "viridis",
            # colorNA = "grey",
            title = "Number of individuals 50+ per 100m2") + 
  tm_shape(eye_fac_sf %>% filter(cat_serv == "Permanent")) + 
  tm_dots(size = 0.05,
          col = "red") +
  tm_shape(eye_fac_sf %>% filter(cat_serv == "Occasional")) + 
  tm_dots(size = 0.05,
          col = "pink") +
  tm_add_legend(type = "symbol",
                col = c("red","pink"),
                labels = c("Permanent cataract surgical facilities",
                           "Occasional cataract surgical facilities")) + 
  tm_layout(legend.outside = T,
            legend.outside.position = "right",
            legend.stack = "vertical",
            legend.frame = T,
            legend.outside.size = 0.5)
constrained_map

# saving 
tmap_save(constrained_map, "data/maps/constrained.pdf")

```

## Repeating the whole process for unconstrained population estimates.

The resulting map has less bare space (indicating areas where no structures could be determined and thus were considered to be uninhabited). The bare space is instead replaced with colors close to purple, indicating 0 or close to 0 inhabitants in thatn 100m2 area. Personally, I think this looks much better than the map that uses constrained estimates.

Note: Inactivating the sections of code responsible for downloading pop data and converting all of them into a single raster. Feel free to reactivate as needed.

```{r repeating the entire process for unconstrained population estimates}

### DOWNLOADING UNCONSTRAINED RASTERS ####

  # creating a list of covariates we want to download
  raster.list <- wpgpListCountryDatasets(ISO3 = "MWI") %>%
    # filtering out the specific covariates we want (constrained estimates of age and sex distribution for those 50+)
    filter(grepl("agesex_", Covariate)) %>%
    filter(grepl("_50", Covariate) | grepl("_55", Covariate) | grepl("_60", Covariate) |
             grepl("_65", Covariate) | grepl("_70", Covariate) | grepl("_75", Covariate) |
             grepl("_80", Covariate)) %>%
    filter(grepl("_2020$", Covariate)) %>%
    dplyr::select(Covariate)
  raster.list
  raster.list.v <- raster.list$Covariate

  # applying it over all the desired covariates
  sapply(X = raster.list.v,  # the covariates we want rasters for
         dir = here("data", "pop_unconstrained"),
         FUN = get.raster,   # the function to apply
         simplify = T)

#### COMBINING RASTERS ####

  # getting a list of all filenames
  list.unconst <- list.files(here("data", "pop_unconstrained"), full.names = T)

  # creating the stack
  popstack.unconst <- stack(list.unconst)

  # giving the stack more intelligible names
  names(popstack.unconst) <- c("f_50","f_55","f_60","f_65","f_70", "f_75", "f_80",
                     "m_50","m_55","m_60","m_65","m_70", "m_75", "m_80")

  # calculating a new raster which is the sum of the population values in all the stacks
  popsum.unconst <- calc(popstack.unconst, fun=sum)

  # renaming the newly created layer
  names(popsum.unconst) <- "total_pop"

  # saving the newly created raster
  writeRaster(popsum.unconst, here("data", "rasters", "popsum_unconstrained.tif"))

#### PLOTTING WITH TMAP ####
  # Opening the tiff file created above (can do this instead of downloading and mergin everything everytime you run the code)
  popsum.unconst <- raster(here("data", "rasters", "popsum_unconstrained.tif"))
  
  # plotting
unconstrained_map <-   tm_shape(malawi) + tm_polygons() +
  tm_shape(popsum.unconst) +
  tm_raster(style = "order",
            palette = "viridis",
            # colorNA = "grey",
            title = "Number of individuals 50+ per 100m2") + 
  tm_shape(eye_fac_sf %>% filter(cat_serv == "Permanent")) + 
  tm_dots(size = 0.05,
          col = "red") +
  tm_shape(eye_fac_sf %>% filter(cat_serv == "Occasional")) + 
  tm_dots(size = 0.05,
          col = "pink") +
  tm_add_legend(type = "symbol",
                col = c("red","pink"),
                labels = c("Permanent cataract surgical facilities",
                           "Occasional cataract surgical facilities")) + 
  tm_layout(legend.outside = T,
            legend.outside.position = "right",
            legend.stack = "vertical",
            legend.frame = T,
            legend.outside.size = 0.5)
unconstrained_map

# saving 
tmap_save(unconstrained_map, "data/maps/unconstrained.pdf")
```

## Alternatively one can turn the raster into point data and plot it.

This provides a lot of overlap in the points and I think either of the above maps are better, but the data in this form may be useful for later analyses. The example below uses constrained population estimates. It is homogenous because there is so much overlap between the points. 

```{r mapping with points}
# converting stack to dataframe
popsum.df <- as.data.frame(rasterToPoints(popsum)) %>%
  rename(total=popsum_constrained)

# checking my work on a random sample
  # ts <- sample_n(popsum.df, 10000) 
  # 
  # ts %>% group_by(x, y) %>% mutate(dups=n()) %>% filter(dups>1) # no duplicates
  # 
  # ts %>% filter(is.na(m_50)) # no data are missing

# Saving the resultant dataframe
# write_csv(popsum.df, here("data","pop_constrained","popsum_constrained_df.csv"))

# converting the dataframe to an SF object
popsum.sf <- popsum.df %>%
  st_as_sf(coords = c("x","y"), # converting to an sf object
           crs = "4326")

# plotting -- note this takes a while
ggplot(data = popsum.sf) +
  geom_sf(data = popsum.sf,
          aes(fill = total, 
              color = total, 
              alpha = 0.05), 
          inherit.aes = F) + 
  scale_fill_viridis() + 
  scale_color_viridis() +
  geom_sf(data = filter(eye_fac_sf, cat_serv == "Permanent"), color = "red") + 
  geom_sf(data = filter(eye_fac_sf, cat_serv == "Occasional"), color = "pink")

```

